Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-26 ms/cover_letter/plosHIVcover_resubm2.tex
various ms/bib/TODO tweaks
still working ...

\response{
\todo{diagram of models}
}

@@

@@

\response{
  added \ldots
}

@@

%\bibliography{virulence}

>>>>

\response{
This is a great idea.  We have added a figure that shows four diagrams (one for each of the central models), with a colour scheme matching the colours used to denote models in later figures.
}

@@

\response{
We agree that our analysis is not by itself enough to assess 
the eco-evolutionary dynamics of HIV in a setting that is realistic
enough to apply in a public policy context. However, we still feel
it represents a valuable contribution to the HIV literature, by
providing guidance to other modelers on both how and why 
eco-evolutionary conclusions can depend on model structure.
}

\response{
We extended our analysis to apply both heterogeneity and vital dynamics
(separately) to all four of the base models (pairform/instswitch with
and without extra-pair contact); we report our findings in the
last paragraph of the Results section and in Figure S4 (Supplementary material 2).
}

\response{
We removed this sentence.
}

@@

@@

\bibliography{virulence}

### 2017-01-26 ms/ms.tex
still working ...

\begin{table}[h!]
\caption{Parameter ranges/values. Parameters with fixed values are kept constant throughout simulations while other parameter values are taken from Latin hypercube samples using ranges specified in the table. Values of $c$ and $\rho$ are doubled from those given by Champredon \etal\ because we keep track of individuals in the model, while they keep track of couples. Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

@@

\caption{{\bf Model contact structures.} 
Schematic representations of models with
explicit contact structure. Top row (a,b) shows models with instantaneous
partnership formation; bottom row (c,d)  shows those where individuals remain
single for some period of time between partnerships. Left column (a,c) 
shows models with within-partnership transmission only; in these models
transmission can only occur within serodiscordant partnerships, and
serodiscordant partnerships can only be formed by dissolution
and reformation of partnerships. Right column (b,d) shows models
that allow extra-pair and uncoupled transmission.
\emph{Solid arrows} represent infection transitions; \emph{dotted arrows}
represent partnership dissolution and formation; \emph{dashed arrows}
show influences on infection rate, including both within-pair
transmission (blue) and extra-pair/uncoupled transmission (red).
In addition to these models, we also use random-mixing and implicit
models with only $S$ and $I$ compartments and heterogeneous models
where each compartment is subdivided by sexual activity rate
(of both partners, in the case of partnership compartments).
}
\label{fig:diagrams}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl\ using baseline parameters. Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution. Note that the initial condition has negligible effect on equilibrium virulence.
(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
(b) Effects of varying initial infectious density $I(0)$.
(c) Effects of varying initial mean virulence $\alpha(0)$.
The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate. 
The eco-evolutionary dynamics are qualitatively similar
across a range of initial conditions (panels b,c).}
\label{fig:panel3}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig2.pdf}
\caption{{\bf Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
Models arranged in order of complexity of contact structure, from 
``hetero'' (most complex) to ``random'' (least complex).
All models were run until $t=4000~\textrm{years}$ using parameters from 1000 Latin hypercube samples to illustrate the range of possible dynamics. Envelopes contain the middle 95\% of trajectories (i.e. we select all points between the 0.025 and 0.975 quantiles for each model at each year), while lines within the envelopes show mean trajectories. Truncated series (up to year 800) are shown here.}
\label{fig:virtraj}
\end{figure}

@@

Finally, incorporating additional realism to the model, i.e.
combining heterogeneity with all four basic contact structures or
allowing for vital dynamics rather than assuming an SIS model,
leads to only small differences in the conclusions stated so far
(\figurename~??). Relative to our baseline SIS assumption, the
effect of adding vital dynamics is to delay the virulence peak
slightly and increase both the peak and equilibrium virulence.
The changes are small, however: across all models, the
maximum increase in time until
the virulence peak is 40 years (for the instswitch+epc model),in 
the peak \Lspvl\ is 0.24 units (instswitch), and in the equilibrium
\Lspvl\ is 0.4 (pairform). The changes in the most realistic
model (pairform+epc with heterogeneity)

>>>>

\begin{table}[h!]
\caption{\textbf{Parameter ranges and values}. Parameters with fixed values are kept constant throughout simulations while other parameter values are taken from Latin hypercube samples using ranges specified in the table. Values of $c$ and $\rho$ are doubled from those given by Champredon \etal\ because we keep track of individuals in the model, while they keep track of couples. Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

@@

\caption{\textbf{Model contact structures.} 
Schematic representations of models with
explicit contact structure. Top row (a,b) shows models with instantaneous
partnership formation; bottom row (c,d)  shows those where individuals remain
single for some period of time between partnerships. Left column (a,c) 
shows models with within-partnership transmission only; in these models
transmission can only occur within serodiscordant partnerships, and
serodiscordant partnerships can only be formed by dissolution
and reformation of partnerships. Right column (b,d) shows models
that allow extra-pair and uncoupled transmission.
\emph{Solid arrows} represent infection transitions; \emph{dotted arrows}
represent partnership dissolution and formation; \emph{dashed arrows}
show influences on infection rate, including both within-pair
transmission (blue) and extra-pair/uncoupled transmission (red).
In addition to these models, we also use random-mixing and implicit
models with only $S$ and $I$ compartments and heterogeneous models
where each compartment is subdivided by sexual activity rate
(of both partners, in the case of partnership compartments).
}
\label{fig:diagrams}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{\textbf{Baseline dynamics.}
Time series of mean population \Lspvl\ using baseline parameters.
(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
(b) Effects of varying initial infectious density $I(0)$.
(c) Effects of varying initial mean virulence $\alpha(0)$.
The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate. The other panels (b,c) show that eco-evolutionary dynamics are qualitatively similar across a range of initial conditions.}
\label{fig:panel3}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig2.pdf}
\caption{\textbf{Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
Panels are arranged in order of complexity of contact structure, from 
``random'' (least complex) to ``hetero'' (most complex).
All models were run until $t=4000~\textrm{years}$ using parameters from 1000 Latin hypercube samples to illustrate the range of possible dynamics. Envelopes contain the middle 95\% of trajectories (i.e. we select all points between the 0.025 and 0.975 quantiles for each model at each year), while lines within the envelopes show mean trajectories. Truncated series (up to year 800) are shown here.}
\label{fig:virtraj}
\end{figure}

@@

Finally, incorporating additional realism to the model, i.e.
combining heterogeneity with all four basic contact structures or
allowing for vital dynamics rather than assuming an SIS model,
leads to only small differences in the conclusions stated so far
(\figurename~??). Relative to our baseline SIS assumption, the
effect of adding vital dynamics is to delay the virulence peak
slightly and increase both the peak and equilibrium virulence.
The changes are small, however: across all models, the
maximum increase in time until
the virulence peak is 40 years (for the instswitch+epc model),in 
the peak \Lspvl\ is 0.24 units (instswitch), and in the equilibrium
\Lspvl\ is 0.4 (pairform). The changes in the most realistic
model (pairform+epc) are considerably smaller: 
an increase of 2 years vs. a decrease of 13 years in the
time to the virulence peak for the
models with vital dynamics and heterogeneity, respectively;
an increase of 0.1 units vs. a decrease of 0.01 units in peak \Lspvl;
and an increase of 0.145 units vs. a decrease of 0.03 units
in equilibrium~\Lspvl. Thus, while we can never rule out the
possibility of some higher-order interaction among epidemiological
phenomena leading to significant changes in our conclusions,
we are reasonably confident that the results reported here are
robust to additional complexities.
